Numerical simulation of the Cordilleran ice sheet through the last glacial cycle

### 2014-06-17 cordillera-cycle.tex
Develop discussion of deglaciation history.

For more than a century, exploration and geological investigation of the
Cordillera have led to a large collection of evidence of the former ice cover
\citep{Jackson.Clague.1991}. This evidence consists of mapped boundaries of the
former ice extent (\citealp[Fig.~1.12]{Clague.1989}; \citealp{Duk-Rodkin.1999};
\citealp{Dyke.2004}), directions of past sliding velocities
(\citealp{Prest.etal.1968}; \citealp[Fig.~2]{Kleman.etal.2010}), and locations
of former
melt-water streams \citep{Margold.etal.2011, Margold.etal.2013}, as well as the
timing of glaciation identified from radiocarbon dating \citep[e.g.,][]
{Clague.1981, Porter.Swanson.1998}, cosmogenic dating \citep[e.g.,][]
{Ward.etal.2007, Menounos.etal.2009, Stroeven.etal.2010, Stroeven.etal.2013},
and the offshore sedimentary record \citep{Cosma.etal.2008, Davies.etal.2011}%
\footnote{\textbf{Arjen, Martin,}, please comment on my choice of references
    in this paragraph. Is it a fair sample of the current state of the field?
    Am I missing some other important works?}.

@@

In the North American Cordillera alike other glaciated regions, the large
majority of the glacial geologic record relates to the last few millennia of
glacier cover, most of the older evidence having been overridden by ice retreat
retreat during the deglaciation.

In our simulations, the timing of the last glacial maximum, and in turn that of
the deglaciation, depends on the palaeo-temperature time-series used to force
the model (Figs.~\ref{fig:timeseries} and ~\ref{fig:deglacseries}). The
6\,km-resolution simulation forced by the EPICA record reaches a maximal
ice volume at 17.3\,kyr, followed by a continuous deglaciation between
about 17 and 11\,kyr. That driven by the GRIP record reaches maximal
volume at 19.1\,kyr, and produces a deglaciation in three steps,
including the first stand-still between 16.6 and 14.5\,kyr, and a late
glacial re-advance between 12.6 and 11.6\,kyr.

\todo{discuss the work that supports YD re-advance.}

Although the predicted timing of glacier retreat differs, its patterns are
relatively consistent between the two simulations (Fig.~\ref{fig:deglac}). In
both runs, the southern half of the modelling domain, including the Puget
Lowland, the Coast and Rocky Mountains and the Interior Plateau of British
Columbia, are completely deglaciated by 12\,kyr, whereas significant
ice cover remains on the Skeena, Selwyn, MacKenzie, Wrangell and Saint-Elias Mountains
in the northern half of the domain. At this stage, a late glacial re-advance
occurs in the GRIP simulation. Both simulations then show an unzipping of the
Liard Lowland, followed by ice retreat towards the Selwyn Mountains and the
Skeena Mountains, which host the last remains of the palaeo-ice sheet.

\todo{discuss existing evidence (Martin).}

On the Interior Plateau of British Columbia, modelled deglaciation consists of
a rapid northwards retreat, leaving a slow-flowing ice lobe on the
plateau, while the surrounding mountain ranges are already deglaciated
(Figs.~\ref{fig:profiles-grip} and~\ref{fig:profiles-epica}). Although this
supports the top-down (?)
deglaciation hypothesis for central British Columbia, modelled, southwards,
last flow directions do not agree with the local lineation patterns
(Fig.~\ref{fig:lastflow}). This could indicate that the lineations are older,
or that the modelled last glacial maximum ice sheet is too extensive to the
east, resulting in modelled deglaciation flow patterns different from those
observed in the geomorphology.

@@

% fig:deglac
\begin{figure*}
  \includegraphics{deglac}
  \caption{Modelled age of the last deglaciation. Areas where the MIS~4 glacial
           advance exceeded the last glacial maximum advanced are marked in
           green. Hatches denote the Younger Dryas re-advance, which is more
           pronounced in the GRIP-driven simulation.}
  \label{fig:deglac}
\end{figure*}

>>>>

For more than a century, exploration and geological investigation of the
Cordillera have led to a large collection of evidence of the former ice cover
\citep{Jackson.Clague.1991}. This evidence consists of mapped boundaries of the
former ice extent (\citealp[Fig.~1.12]{Clague.1989}; \citealp{Duk-Rodkin.1999};
\citealp{Dyke.2004}), directions of past sliding velocities
(\citealp{Prest.etal.1968}; \citealp[Fig.~2]{Kleman.etal.2010}), and locations
of former
melt-water streams \citep{Margold.etal.2011, Margold.etal.2013a}, as well as the
timing of glaciation identified from radiocarbon dating \citep[e.g.,][]
{Clague.1981, Porter.Swanson.1998}, cosmogenic dating \citep[e.g.,][]
{Ward.etal.2007, Menounos.etal.2008, Stroeven.etal.2010, Stroeven.etal.2013},
and the offshore sedimentary record \citep{Cosma.etal.2008, Davies.etal.2011}%
\footnote{\textbf{Arjen, Martin,}, please comment on my choice of references
    in this paragraph. Is it a fair sample of the current state of the field?
    Am I missing some other important works?}.

@@

In the North American Cordillera alike other glaciated regions, the large
majority of the glacial geologic record relates to the last few millennia of
glacier cover, most of the older evidence having been overridden by ice retreat
retreat during the deglaciation \footnote{\textbf{Johan,} can we refer to one
of your papers here?}.

In our simulations, the timing of the last glacial maximum, and in turn that of
the deglaciation, depends on the palaeo-temperature time-series used to force
the model (Fig.~\ref{fig:timeseries}). The
6\,km-resolution simulation forced by the EPICA record reaches a maximal
ice volume at 17.3\,kyr, followed by a continuous deglaciation between
about 17 and 11\,kyr (Fig.~\ref{fig:deglacseries}, red dotted curve).
That driven by the GRIP record reaches maximal
volume at 19.1\,kyr, and produces a deglaciation in three steps,
including the first stand-still between 16.6 and 14.5\,kyr, and a late
glacial re-advance between 12.6 and 11.6\,kyr (Fig.~\ref{fig:deglacseries},
blue dotted curve).

The possibly of a widespread late-glacial re-advance of decaying glaciers of
the North American Cordilleran has been debated for long. On the southern tip
of the Coast Mountains, radiocarbon-dated frontal moraines indicate multiple
stages of glacial re-advance in the Fraser and Squamish valleys, of which
one corresponds to the Younger Dryas chronozone \citep{Clague.etal.1997,
Friele.Clague.2002, Friele.Clague.2002a, Kovanen.2002,
Kovanen.Easterbrook.2002}. Additional evidence for a late-glacial re-advance
was found in the Finlay River area, northern Rocky Mountains, where
sharp-crested moraines show an abundant remobilization of late-glacial
sediments by alpine glaciers in interaction with the main body of the decaying
ice sheet \citep{Lakeman.etal.2008}. Although further work is needed to
constrain the timing of the North American Cordillera late-glacial re-advance,
to seize its extent and identify potential climatic triggers
\citep{Menounos.etal.2008}, it is interesting to note that the simulation
driven by the GRIP record produces a late-glacial re-advance in the Coast
Mountains, Rocky Mountains and Finlay River area, corresponding to where it
has been identified in the geologic record (Fig.~\ref{fig:deglac}, left panel),
whereas the EPICA-driven simulation produces a nearly-continuous deglaciation
with very restrained glacial re-advance (Fig.~\ref{fig:deglac}, right panel).

Although the predicted timing of glacier retreat differs between the two
simulations, its patterns are relatively consistent (Fig.~\ref{fig:deglac}). In
both runs, the southern half of the modelling domain, including the Puget
Lowland, the Coast and Rocky Mountains and the Interior Plateau of British
Columbia, are completely deglaciated by 12\,kyr, whereas significant ice cover
remains on the Skeena, Selwyn, MacKenzie, Wrangell and Saint-Elias Mountains
in the northern half of the domain. After 12\,kyr, deglaciation proceeds
through an opening of the Liard lowland with radial margin retreat towards the
surrounding mountain ranges and inward flow towards the depression, in full
compatibility with the regional melt water record \citep{Margold.etal.2013}.
Remaining ice continues its decay by retreating towards the Selwyn Mountains
and, then, the Skeena Mountains. The last remains of the palaeo-ice sheet
vanish in the Skeena Mountains at around 10.6\,kyr in the EPICA simulation
versus 9.4\,kyr in the GRIP simulation.

Once more, snapshots of modelled deglacial ice sheet configurations
(Figs.~\ref{fig:deglacshots-grip} and~\ref{fig:deglacshots-epica}) provide a
poor medium for direct comparison with geomorphological evidence of the ice
retreat. Here, we compute a map of de-glacial, basal flow direction immediately
preceding exposure of ice-free land or cessation of sliding upon the approach
of the retreating margin (Fig.~\ref{fig:lastflow}). On the northern and
southern front of the ice sheet and in the Liard Lowland, the resulting map
show similarities with patterns of glacial lineations
(\citealp{Prest.etal.1968}; \citealp[Fig.~1.12]{Clague.1989};
\citealp[Fig.~2]{Kleman.etal.2010}; \citep[Fig.~2]{Margold.etal.2013}).
However, such is not the case on the Interior Plateau of British Columbia,
where the model produce no deglacial flow directions (Fig.~\ref{fig:lastflow}),
whereas glacial lineation maps indicate eastward flow.

On the one hand, this could indicate that the model performs poorly in this
region, perhaps because the modelled last glacial maximum ice sheet is too
thick, or too extensive to the west. Because feedback mechanisms between ice
sheet topography and regional climate such as wind redistribution, orographic
precipitation and latent warming of moisture-depleted air are absent in our
model, we regard this possibility as fully probable \citep{Seguinot.etal.2013}.

On the other hand, in both simulations driven by the GRIP and EPICA records,
the deglaciation of central British Columbia consists of a rapid northwards
retreat associated with southwards-flowing, non-sliding ice lobes on the
Interior Plateau, while the surrounding mountain ranges are already ice-free
(Figs.~\ref{fig:profiles-grip} and~\ref{fig:profiles-epica}). Thus the model
suggests that glacial lineations on the Interior Plateau of British Columbia
may be of older age than the last glacial maximum, and may have remained intact
throughout the deglaciation.

@@

% fig:deglac
\begin{figure*}
  \includegraphics{deglac}
  \caption{Modelled age of the last deglaciation. Areas where the MIS~4 glacial
           advance exceeded the last glacial maximum advanced are marked in
           green. Hatches denote re-advance of mountain-centred ice caps and
           and the decaying ice sheet between 14 and 10\,kyr., which is more
           pronounced in the GRIP-driven simulation.}
  \label{fig:deglac}
\end{figure*}
